Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2017-06-19 2017-Xudong-SMT/submit-ICTAC2017.tex
Complete a part of the TODO.
Add the CCS example and result explanation.
Move the 4.1 and 4.2 to the Section 5. Stress that the real implementation has more stuff than the sketch.

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.
In this paper, we use an example inspired from Lotos, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking.

@@

It should be easy to see that this pNet system encodes properly the
behavioral semantics of the corresponding Lotos expression. We remark
that it is built in a systematic structural way, one pNet node
encoding each Lotos operator. It is parameterized both at the level of
actions variables (occurring within the controllers and the
synchronization vectors), and at the level of process parameters
(holes).

@@

\subsection{Presentation of algebra}
One of the important goal of the Open pNet framework is to provide us
with a model able to express the semantics of many different algebras
or languages. In particular, as we have seen in section
\ref{section:pnets}, the \emph{term algebra} used to express the
action expressions and the data in their parameters is left open in
the pNet formalization.

So, for a given language, we now need to specify its data and action
domains, giving them an abstract syntax (sorts, constructors, operators,
and predicates). We call this a \emph{Presentation}.
We also define a concrete syntax, that will be used for
pretty-printing, but also for translating the presentation, and the
predicates, into Z3 syntax (we shall detail this translation in
section \ref{section:pruning}).

An algebra presentation contains:
\begin{itemize}
	\item Sorts: Constants sets of the algebra or types of the data
          and actions. In every presentation, sorts for integer and
          boolean are implicitly declared.  There \emph{must} be one sort for actions, and eventually 
several others for the types of data parameters. In the LOTOS example,
we have the $Action$ sort: $\OTvar{Action = } \{ l, \delta, r, p, q, acc \} .$

@@

\TODO{To be move / replace by Xudong}
\subsection{Computing and using open automata}
In this section we present an algorithm to construct the open
automaton representing the behavior of an open pNet, and show
under which conditions this automaton is finite.

\begin{alg}[behavioral semantics of open pNets: Sketch]
This is a standard residual algorithm over a set of open-automaton
states, but where transitions are open transitions
constructively ``proven'' by deduction trees.

1) Start with a set of unexplored states containing the initial state
of the automaton, and an empty set of explored states.

2) While there are unexplored states:

2a) pick one state from the unexplored set and add it to the explored
set. From this state
build all possible deduction trees by application of the structural
rules Tr1 and Tr2, using all applicable combinations
of synchronization vectors.

2b) For each deduction trees, extract the resulting
open-transition.
Its predicate is checked (using Z3) for  satisfiability. If it is
unsatisfiable, the resulting OT is discarded. This will minimize the
number of resulting transitions.

2c) For each open-transition,  add the transition in the outgoing transitions of
the current state, and add the
resulting state in the unexplored set if it is not already in the
explored set.

3) Finally a reachability check could potentially
prune the state space.

\end{alg}

@@

\subsection{Combining and Matching using TR2}
\TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}

@@

@@

\subsubsection{Result of CCS example}
\QIN{
also can present all the possible behavior of that CCS formula with 16 open
transitions in Fig. ~\ref{schema:ccs-result}, here we omitted the details of open transitions. 
In this example, there are 32 more transitions detected unsatisfiable by Z3.
}
\begin{figure}[h]
  \centerline{\includegraphics[width=5cm]{XFIG/CCSresult}}
\caption{The open automaton of the CCS parallel formula}  \label{schema:ccs-result}
\end{figure}

>>>>

Several publications \cite{HMZ:PDP15,henrio:Forte2016} already have introduced
many examples of pNets, encoding
operators of various classical process algebras, or more complex
synchronization structures in distributed or parallel languages.

In this paper, we use an example inspired from LOTOS, to illustrate
the main features of pNets and the way we deal with them in the symbolic
semantics, and in the satisfiability checking. 
\QIN{
Meanwhile, 
the formula using operators from CCS can also be encoded into a pNet and 
conducted with the same things.
}

\subsection{LOTOS Formula Example}

@@

\subsection{CCS Parallel Example}

\begin{figure}[h]
  \centerline{\includegraphics[width=10cm]{XFIG/CCS1.eps}}
  \caption{pNet encodings for the CCS expression}  \label{schema:ccs-pnet}
\end{figure}

\QIN{
This CCS expression composing two operators, parallel and prefix. 
It is easy to recognize that the pNets $PN2$ and $PN3$ are similar to 
the sub-pNet in the LOTOS example in their structure and synchronization vectors. 
That is because they are the operators conduct a similar behavior from different algebras.
}

\QIN{
To present the parallel property, we have defined a set of synchronization vectors composing 
two subnets. We use the notation from the CCS to present the emit and receive action. 
The input variables $v_1$, $v_2$ are the data transmitting between two components, 
showing its value-passing behavior.
}

It should be easy to see that this pNet system encodes properly the
behavioral semantics of the corresponding algebra expression. We remark
that it is built in a systematic structural way, one pNet node
encoding each operator. It is parameterized both at the level of
actions variables (occurring within the controllers and the
synchronization vectors), and at the level of process parameters
(holes).

@@

%\subsection{Presentation of algebra}
%One of the important goal of the Open pNet framework is to provide us
%with a model able to express the semantics of many different algebras
%or languages. In particular, as we have seen in section
%\ref{section:pnets}, the \emph{term algebra} used to express the
%action expressions and the data in their parameters is left open in
%the pNet formalization.
%
%So, for a given language, we now need to specify its data and action
%domains, giving them an abstract syntax (sorts, constructors, operators,
%and predicates). We call this a \emph{Presentation}.
%We also define a concrete syntax, that will be used for
%pretty-printing, but also for translating the presentation, and the
%predicates, into Z3 syntax (we shall detail this translation in
%section \ref{section:pruning}).
%
%An algebra presentation contains:
%\begin{itemize}
%	\item Sorts: Constants sets of the algebra or types of the data
%          and actions. In every presentation, sorts for integer and
%          boolean are implicitly declared.  There \emph{must} be one sort for actions, and eventually 
%several others for the types of data parameters. In the LOTOS example,
%we have the $Action$ sort: $\OTvar{Action = } \{ l, \delta, r, p, q, acc \} .$
%
%	\item Operators: Operators used to construct action
%          expressions (including their data parameters); they are
%          constructors of data and action sorts, but also predicates
%          over these objects. For instance, in our LOTOS example, we
%          could have two constructor functions ``$DELTA$'' and ``$ACC$''
%          to construct the action expression $\delta(x)$ and $acc(x)$, declared as:
%          $\OTvar{DELTA:}\OTvar{< Data >} \rightarrow \OTvar{Action .}$
%        \item Constants: The satisfiability solver need to know which
%          actions in the pLTS behaviors are constants (like the $l$, $\delta$, 
%          $r$ of the controllers in running example). 
%          For a given pNet system, a sort $Action$ will include all
%          these constants.
%\end{itemize}

@@

\TODO{To be move / replace by Xudong}
%\subsection{Computing and using open automata}
%In this section we present an algorithm to construct the open
%automaton representing the behavior of an open pNet, and show
%under which conditions this automaton is finite.
%
%\begin{alg}[behavioral semantics of open pNets: Sketch]
%This is a standard residual algorithm over a set of open-automaton
%states, but where transitions are open transitions
%constructively ``proven'' by deduction trees.
%
%1) Start with a set of unexplored states containing the initial state
%of the automaton, and an empty set of explored states.
%
%2) While there are unexplored states:
%
%2a) pick one state from the unexplored set and add it to the explored
%set. From this state
%build all possible deduction trees by application of the structural
%rules Tr1 and Tr2, using all applicable combinations
%of synchronization vectors.
%
%2b) For each deduction trees, extract the resulting
%open-transition.
%Its predicate is checked (using Z3) for  satisfiability. If it is
%unsatisfiable, the resulting OT is discarded. This will minimize the
%number of resulting transitions. 
%
%2c) For each open-transition,  add the transition in the outgoing transitions of
%the current state, and add the
%resulting state in the unexplored set if it is not already in the
%explored set.
%
%3) Finally a reachability check could potentially
%prune the state space.
%
%\end{alg}

@@

\subsection{Combining and Matching using TR2}
\TODO{Xudong: move here a (new version) of the alogorithm loplevel sketch}
\QIN{
In previous work, we already have a sketch for computing the open automaton.
}

\QIN{
1) Start with the initial state of the automaton. 
}

\QIN{
2) Apply Tr1 and Tr2 to deduct all the possible open transitions. Use SMT solver to prune the 
unsatisfiable transitions. Recursively conduct this step with the target state of the transitions 
until all the global states are explored.
}

\QIN{
3) Finally a reachability check could potentially prune the state space.
}

\QIN{
The detail about the application of the Tr1 and Tr2 have not considered in the sketch, but it is an  
important part in the implementation of open transition generation. Each time we apply Tr1, there is not too much predicate terms, so do the other parameters, needed 
to be generated. When it comes to applying Tr2, the computation becomes complicated because it doing some composition of the subnets.
}

@@

\paragraph{Presentation of algebra}
\QIN{
Open pNet framework provides us
with a model able to express the semantics of many different algebras
or languages. In particular, as we have seen in section
\ref{section:pnets}, the \emph{term algebra} used to express the
action expressions and the data in their parameters is left open in
the pNet formalization.
}

\QIN{
So, for a given language, we now need to specify its data and action
domains, giving them an abstract syntax (sorts, constructors, operators,
and predicates). We call this a \emph{Presentation}.
We also define a concrete syntax, that will be used for
pretty-printing, but also for translating the presentation, and the
predicates, into Z3 syntax.
}

\QIN{
An algebra presentation contains:
}

\begin{itemize}
	\item Sorts: Constants sets of the algebra or types of the data
          and actions. In every presentation, sorts for integer and
          boolean are implicitly declared.  There \emph{must} be one sort for actions, and eventually 
several others for the types of data parameters. In the LOTOS example,
we have the $Action$ sort: $\OTvar{Action = } \{ l, \delta, r, p, q, acc \} .$

@@

\subsubsection{Result of CCS example}
\QIN{
also can present all the possible behavior of that CCS formula with 16 open
transitions in Fig. ~\ref{schema:ccs-result}, here we omitted the details of open transitions. 
In this example, there are 32 more transitions detected unsatisfiable by Z3.
}

\QIN{
Here we only show one of the result transitions. It presents the value-passing between $a.P$ and $b.Q$ and they are synchronized then
give out an silent action $\tau$, process $P$ and $Q$ should not work in this case.
}

\begin{figure}[h]
  \centerline{\includegraphics[width=5cm]{XFIG/CCSresult}}

  $\begin{array}{@{}r@{}l@{}}
    ot_1  = &\openrule{\{0 \xrightarrow{a}_{C_1} 1, 0 \xrightarrow{b}_{C_2} 1\}, ~~a=!ch_{11}(v_{11}) \!\land\! b=?ch_{11}(v_{11}) \!\land\! a\neq l \!\land\! b\neq r\!\land\! v_1=\tau}
    {\ostate{00} \xrightarrow{v_1} \ostate{11}}\\
  \end{array}$

\caption{The open automaton of the CCS parallel formula}  \label{schema:ccs-result}
\end{figure}
